# 1

Load packages, rename variables to be more descriptive, and recode the dummy
variables as logical.

```{r}
library(car)
library(dplyr)
library(purrr)
library(readr)
library(sandwich)
library(lmtest)

commute_binary = 
    read_csv("commute_binary.csv") %>%
    rename(
        individual_id = id,
        commute_mode = mode,
        drive_time_minutes = time.car,
        drive_cost_dollars = cost.car,
        bus_time_minutes = time.bus,
        bus_cost_dollars = cost.bus,
        gas_price_dollars = price_gas,
        daily_snow_inches = snowfall,
        age_years = age,
        income_thousand_dollas = income
    ) %>%
    mutate(
        construction = as.logical(construction),
        bus_detour = as.logical(bus_detour),
    )
```

## a

Add a variable that is `TRUE` if someone drives, and `FALSE` otherwise.

```{r}
commute_binary = 
    commute_binary %>%
    mutate(
        drives = commute_mode == "car"
    )
```

Calculate the probability model.

```{r}
probability_model = lm(
    drives ~ drive_cost_dollars + drive_time_minutes + bus_time_minutes,
    commute_binary
)
```

### i

Here are the estimated coefficients.

```{r}
coeftest(probability_model)
```

Here are the heteroskedastic-robust standard errors.

```{r}
vcovHC(probability_model)
```

### ii

Calculate the predicted probabilties, and see how many are infeasible, that is,
not between 0 and 1

```{r}
probability_model_prediction = predict(probability_model, commute_binary)
sum(!(0 <= probability_model_prediction & probability_model_prediction <= 1))
```

### iii

```{r}
linearHypothesis(probability_model, "drive_time_minutes = bus_time_minutes")
```

# 2

## a

### i

Calculate the logistic model.

```{r}
logistic_model =
    glm(
        drives ~ drive_cost_dollars + drive_time_minutes + bus_time_minutes,
        commute_binary,
        family = "binomial"
    )
summary(logistic_model)
```

### ii

For convenience, define

$$ Y_n = β_0 + β_1 C_{nc} + β_2 T_{nc} + β_3 T_{nb} $$

Then

$$ ln \left( \frac{P_n}{1 - P_n} \right) = Y_n $$

Solve for $P$.

$$ \frac{P_n}{1 - P_n} = e^{Y_n} $$
$$ P_n = e^{Y_n}(1 - P_n) $$
$$ P_n = e^{Y_n} - P_n e^{Y_n} $$
$$ P_n + P_n e^{Y_n} = e^{Y_n} $$
$$ P_n (1 + e^{Y_n}) = e^{Y_n} $$
$$ P_n = \frac{e^{Y_n}}{1 + e^{Y_n}} $$

First, find $\frac{d P_n}{d Y_n}$

$$ \frac{d P_n}{d Y_n} = \frac{(1 + e^{Y_n}) e^{Y_n} - (e^{Y_n})^2} {(1 + e^{Y_n})^2} $$
$$ \frac{d P_n}{d Y_n} = \frac{e^{Y_n} + (e^{Y_n})^2 - (e^{Y_n})^2} {(1 + e^{Y_n})^2} $$
$$ \frac{d P_n}{d Y_n} = \frac{e^{Y_n} + (e^{Y_n})^2 - (e^{Y_n})^2} {(1 + e^{Y_n})^2} $$
$$ \frac{d P_n}{d Y_n} = \frac{e^{Y_n}} {(1 + e^{Y_n})^2} $$

In R,

```{r}
expY = exp(predict(logistic_model, commute_binary))
dpdY = expY / (1 + expY)^2
dPdY_summary = summary(dpdY)
```

By the chain rule,

$$ \frac{d P_n}{d C_{nc}} = \frac{d P_n}{d Y_n} \frac{d Y_n}{d C_{nc}} $$
$$ \frac{d P_n}{d C_{nc}} = β_1 \frac{d P_n}{d Y_n} $$

The formulas for the other marginal effects are similar. Because these are
linear transformations, we can just multiply the summary statistics of
$\frac{d P_n}{d Y_n}$ by the respective coefficients.

Define a quick function to remove the intercept from coefficients.

```{r}
remove_intercept = function(coefficients) {
    coefficients[names(coefficients) != "(Intercept)"]
}
```

```{r}
logistic_coefficients = coef(logistic_model)
lapply(
    remove_intercept(logistic_coefficients),
    function(coefficient) {
        dPdY_summary * coefficient
    }
)
```

The marginal effects of the probability model are constant, and just equal to
the coefficients.

```{r}
remove_intercept(coef(probability_model))
```

These are similar to the average marginal effects from the logisitic model
above.

### iii

Start with the model and take the total derivative:

$$ d Y = β_1 d C_c + β_2 d T_c + β_3 d T_b $$

Divide through by $-β_1$. Since $β_1$ is negative, $-β_1$ is positive, so
we don't have to flip the inequality.

$$ - \frac{β_0}{β_1} - C_c - \frac{β_2}{β_1} T_c - \frac{β_3}{β_1} T_b > 0 $$

We can reconcile these two models by setting

$$ C_b = - \frac{β_0}{β_1} $$
$$ k_c = - \frac{β_2}{β_1} $$
$$ k_b = \frac{β_3}{β_1} $$

Finally, we will need to convert the units of $k_c$ and $k_b$. To convert to
dollars per hour, multiply by 60, because

$$ \frac{\text{dollars}}{\text{hour}} = \frac{\text{dollars}}{\text{minute}} * \frac{\text{60 minutes}}{\text{hour}} $$

The cost of spending time driving, in dollars per hour, is

```{r}
60 * logistic_coefficients[["drive_time_minutes"]] /
  logistic_coefficients[["drive_cost_dollars"]]
```

The cost of spending time riding the bus, in dollars per hour, is

```{r}
- 60 * logistic_coefficients[["bus_time_minutes"]] /
  logistic_coefficients[["drive_cost_dollars"]]
```

# b

Add a variable for the drive-cost income ratio:

```{r}
commute_binary = 
    commute_binary %>%
    mutate(
        drive_cost_income_ratio = 
            drive_cost_dollars / (income_thousand_dollas * 1000)
    )
```

```{r}
income_model =
    glm(
        drives ~ drive_cost_income_ratio + drive_time_minutes + bus_time_minutes,
        commute_binary,
        family = "binomial"
    )
summary(income_model)
```








