---
title: Problem set 1
output: pdf_document
author: "Brandon Taylor"
---

```{r}
# prerequisites for other packages
library(carData)
library(zoo, warn.conflicts = FALSE)

library(car)
library(broom)
library(dplyr, warn.conflicts = FALSE)
library(lmtest)
library(pander)
library(readr)
library(sandwich)
library(tidyr)
```

# 1

## a

### i

```{r}
commute_binary = read_csv("commute_binary.csv", show_col_types = FALSE)

commute_binary_long = 
    commute_binary %>%
    select(id,
        `Chosen commute mode` = mode,
        `Drive cost in $` = cost.car,
        `Drive time in minutes` = time.car,
        `Bus time in minutes` = time.bus
    ) %>%
    pivot_longer(
        c(
            `Drive cost in $`,
            `Drive time in minutes`,
            `Bus time in minutes`
        ),
        names_to = "Variable"
    )

get_median_and_mean = function(data) {
    data %>%
    summarize(
        `Median value` = median(value),
        `Mean value` = mean(value),
        .groups = "drop"
    )
}

commute_binary_long %>%
    group_by(Variable) %>%
    get_median_and_mean() %>%
    pander()
```

For all three variables, the mean is higher than the median, suggesting right skewed distributions.

## b

### ii

```{r}
commute_binary_long %>%
    group_by(`Chosen commute mode`, Variable) %>%
    get_median_and_mean() %>%
    pander()
```

Those who choose to drive have lower mean drive times and costs.
These students might choose to drive because of this.
Counterintuitively, those who choose to take the bus have higher mean bus times.
The medians follow the same pattern as the means, with the exception that the median bus time is the same for both groups.

# 2

## a

Calculate the linear probability model.

```{r}
model = lm(I(mode == "car") ~ cost.car + time.car + time.bus, commute_binary)
```

### i

Here are the coefficients and heteroskedastic-robust standard errors.

```{r}
coeftest(model, vcov. = vcovHC(model)) %>%
    tidy() %>%
    pander()
```

All coefficients are significantly different from 0 at the $\alpha = 0.2\%$ level.

```{r}
coefficients = coef(model)
```

The coefficients mean the following:

- When drive time increases by 1 minute, the probabilility someone will drive decreases by `r pander(-coefficients[["time.car"]] * 100)`% on avarage
- When drive cost increases by $1, the probability someone will drive decreases by `r pander(-coefficients[["cost.car"]] * 100)`% on avarage
- When bus time increases by 1 minute, the probability someone will drive increases by `r pander(coefficients[["time.bus"]] * 100)`% on average

### ii

```{r}
predictions = predict(model)
number_of_infeasible_probabilities = sum(!(0 <= predictions & predictions <= 1))
```

Using this model, we predict `r pander(number_of_infeasible_probabilities)` students have infeasible probabilites of driving.
This seems like a small number out of `r nrow(commute_binary)` students, so I am not overly worried about using a linear probability model.

### iii

```{r}
linearHypothesis(model, "time.car = -time.bus") %>%
    pander()
```

The coefficients for drive time and bus time are significantly different.
The coefficient for drive time is higher in absolute value than the coefficient for bus time.
This might be because students enjoy time on the bus more than time in the car, because students can study or sleep on the bus.
If a delay were to increase equally the time to drive and the time to take the bus, I would expect the proportion of drivers to decrease, because people would prefer to spend the extra time on the bus rather than in a car.
