---
title: Problem Set 2
author: "Brandon Taylor"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r}
library(dfidx, warn.conflicts = FALSE)
library(gmm)
library(mlogit)
library(pander)
library(purrr)
library(readr)
library(tidyr)

# load last so we use filter from dplyr
library(dplyr, warn.conflicts = FALSE)
```

# 1

```{r}
commute_binary <-
    read_csv("commute_binary.csv", show_col_types = FALSE) %>%
    mutate(drove = mode == "car")
```

## a

### i

$$ V_{nc} = β_0 + β_1 C_{nc} + β_2 T_{nc} $$
$$ U_{nc} = V_{nc} + \epsilon_{nc} $$

$$ V_{nb} = β_1 C_{nb} + β_{tb} T_{nb} $$
$$ U_{nb} = V_{nb} + \epsilon_{nb} $$

where, in addition to the variables defined in the prompt,

| Variable | Value |
| - | - |
| $β_{tb}$ | Coefficient for bus time |
| $V_{nc}$ | Representative utility of student $n$ if they drive |
| $V_{nb}$ | Representative utility of student $n$ if they take the bus |
| $U_{nc}$ | Utility of student $n$ if they drive |
| $U_{nb}$ | Utility of student $n$ if they take the bus |
| $\epsilon_{nc}$ | Influence of unobserved characteristics of student $n$ if they drive |
| $\epsilon_{nb}$ | Influence of unobserved characteristics of student $n$ if they take the bus |

For all students, normalize utility such that:

- The intercept of representative utility of students if they take the bus is 0.
- $var \left( \epsilon_{nc} \right) = var \left( \epsilon_{nb} \right) = \frac{{\pi}^2}{6}$

Because this normalization is arbitrary, we can't compare utility between students.

$C_{nb}$ is always 0.

```{r}
all(commute_binary$cost.bus == 0) %>% pander()
```

Let $β_3 = - β_{tb}$.

Then,

$$ V_{nc} - V_{nb} = β_0 + β_1 C_{nc} + β_2 T_{nc} + β_3 T_{nb} $$

```{r}
binary_model <- glm(
    drove ~ cost.car + time.car + time.bus,
    commute_binary,
    family = "binomial"
)

pander(summary(binary_model))
```

```{r}
binary_coefficients <- coef(binary_model)
pander(binary_coefficients[["(Intercept)"]])
```

When drive cost increases by $1, a student's utility if they *drive* increases by

```{r}
binary_cost_coefficient <- binary_coefficients[["cost.car"]]
pander(binary_cost_coefficient)
```

When drive time increases by 1 minute, a student's utility if they *drive* increases by

```{r}
binary_car_time_coefficient <- binary_coefficients[["time.car"]]
pander(binary_car_time_coefficient)
```

When bus time increases by 1 minute, a student's utility if they take the *bus* increases by

```{r}
binary_bus_time_coefficient <- binary_coefficients[["time.bus"]]
pander(-binary_bus_time_coefficient)
```

This is negative because $β_3 = - β_{tb}$

### ii

For marginal effects,

$$ \frac{ ∂P_{ni} }{ ∂z_{ni} } = β_z P_{ni} \left( 1 - P_{ni} \right) $$

For cross-marginal effects,

$$ \frac{ ∂P_{ni} }{ ∂z_{nj} } = - β_z P_{ni} P_{nj} $$

There is one cross-marginal coefficient.
$β_{tb} = - β_3$, and since there are only two alternatives, $P_{nj} = 1 - P_{ni}$.
The negatives cancel out, so I can use the same formula for all three coefficients.

```{r}
binary_predictions <- predict(binary_model, type = "response")
prediction_products <- binary_predictions * (1 - binary_predictions)

get_quartiles_and_mean <- function(vector) {
    # quartiles by default
    # turn the quantiles into a tibble and add the mean
    quantile(vector) %>%
        as.list() %>%
        as_tibble() %>%
        mutate(Mean = mean(vector))
}

tibble(
    Coefficient = c(
        "Car cost in dollars",
        "Drive time in minutes",
        "Bus time in minutes"
    ),
    symbol = c("cost.car", "time.car", "time.bus")
) %>%
    group_by(Coefficient) %>%
    summarize(
        get_quartiles_and_mean(
            prediction_products * binary_coefficients[[symbol]]
        ),
        .groups = "drop"
    ) %>%
    pander()
```

The mean marginal effects are very close to the estimates from problem set 1.

### b

By the implicit function theorem,

$$ \frac{∂C_{nc}}{∂T_{nc}} = -\frac{β_2}{β_1} $$
$$ \frac{∂C_{nb}}{∂T_{nb}} = \frac{β_3}{β_1} $$

Both are in dollars per minute.
To get dollars per hour, I will multiply by 60.

When drive time goes up by an hour, and drive costs go up by $- 60 \frac{β_2}{β_1}$ dollars, the student remains indifferent, because $V_{nc}$ is fixed.
The student must value the hour at $- 60 \frac{β_2}{β_1}$ dollars, to balance out the extra cost they have to pay.
I can use a similar argument for bus time.

So, in dollars, a student values an hour in the car at 

```{r}
pander(-60 * binary_car_time_coefficient / binary_cost_coefficient)
```

In dollars, a student values an hour on the bus at

```{r}
pander(60 * binary_bus_time_coefficient / binary_cost_coefficient)
```

## b

### i

The representative utility functions are now:

$$ V_{nc} = β_0 + β_1 \frac{C_{nc}}{I_n} + β_2 T_{nc} $$
$$ V_{nb} = β_1 \frac{C_{nb}}{I_n} - β_3 T_{nb} $$

Costs are in dollars, but income is in thousands of dollars, so 1 unit of the cost-income ratio represents 0.001.

```{r}
income_model <- glm(
    drove ~ I(cost.car / income) + time.car + time.bus,
    commute_binary,
    family = "binomial"
)

summary(income_model) %>%
    pander()
```

When the drive cost-income ratio and drive time are both 0, a student's representative utility if they drive is

```{r}
income_coefficients <- coef(income_model)
pander(income_coefficients[["(Intercept)"]])
```

When the drive cost-income ratio increases by 1 unit (0.001), a student's utility if they drive increases by

```{r}
cost_income_ratio_coefficient <- income_coefficients[["I(cost.car/income)"]]
pander(cost_income_ratio_coefficient)
```

When drive time increases by 1 minute, a student's utility if they drive increases by

```{r}
income_car_time_coefficient <- income_coefficients[["time.car"]]
pander(income_car_time_coefficient)
```

When bus time increases by 1 minute, a student's utility if they take the bus increases by

```{r}
income_bus_time_coefficient <- income_coefficients[["time.bus"]]
pander(-income_bus_time_coefficient)
```

This is negative because $β_3 = - β_{tb}$

### ii

$$ \frac{∂V_{nc}}{∂C_{nc}} = \frac{β_1}{I_n} $$

Costs subtract from income, so this is the negative marginal utility of a dollar.
However, if one only uses utility to represent individual choices, utility is not interpersonally comparable.

```{r}
incomes_in_thousands <- c(15, 25, 35)
# format the names as dollars
names(incomes_in_thousands) <- paste0("$", incomes_in_thousands, ",000")
marginal_utilities <- -cost_income_ratio_coefficient / incomes_in_thousands
pander(marginal_utilities)
```

By the implicit function theorem:

$$ \frac{∂C_{nc}}{∂T_{nc}} = - \frac{β_2 I_n}{β_1} $$
$$ \frac{∂C_{nb}}{∂T_{nb}} = \frac{β_3 I_n}{β_1} $$

Both are in dollars per minute.
To get dollars per hour, I will multiply by 60.

In dollars, at these incomes, students value an hour in the car at 

```{r}
pander(
    -60 * 
    income_car_time_coefficient *
    incomes_in_thousands / cost_income_ratio_coefficient
)
```

In dollars, at these incomes, students value an hour on the bus at 

```{r}
pander(
    60 * 
    income_bus_time_coefficient * 
    incomes_in_thousands / cost_income_ratio_coefficient
)
```

# 2

## a

### i

In addition to the variables in the prompt, define:

| Variable | Value |
| - | - |
| $\epsilon_{nj}$ | Influence of unobserved characteristics of student $n$ if they choose alternative $j$ |

For all students, normalize utility such that:

- The intercept of representative utility of students if they bike is 0.
- $∀ j \left( var \left( \epsilon_{nj} \right) = \frac{{\pi}^2}{6} \right)$

Because both normalizations are arbitrary, we can't compare utility, or even utility differences, between students.

```{r}
commute_multinomial <-
    read_csv("commute_multinomial.csv", show_col_types = FALSE)

add_indices <- function(data) {
    dfidx(data, shape = "wide", varying = 3:10, choice = "mode")
}

indexed <- add_indices(commute_multinomial)

# the first part is variables that vary by alternative
# but have the same coefficient for all alternatives
# the second part is variables that don't vary by alternative
# 1 stands for the intercept
# the third part is variables that vary by alternative
# and a have different coefficients for each alternative
multinomial_formula <- mode ~ cost | 1 | time
multinomial_model <- mlogit(multinomial_formula, indexed)
summary(multinomial_model)
```

```{r}
multinomial_coefficients <- coef(multinomial_model)

# these are easier for me to interpret in table form
get_coefficients_table <- function(multinomial_coefficients) {
    multinomial_coefficients %>%
        # convert to tibble
        as.list() %>%
        as_tibble() %>%
        # stack coefficient-alternative combos
        pivot_longer(
            c(
                `(Intercept):bus`,
                `(Intercept):car`,
                `(Intercept):walk`,
                `time:bike`,
                `time:bus`,
                `time:car`,
                `time:walk`
            ),
            names_to = c("coefficient", "Alternative"),
            names_sep = "\\:",
            values_to = "value"
        ) %>%
        # put coefficients in columns
        pivot_wider(names_from = coefficient, values_from = value) %>%
        # reorder to match model
        select(Alternative, `(Intercept)`, cost, time)
}

pander(get_coefficients_table(multinomial_coefficients))
```

- When the time and cost of the alternative are both zero, a student's representative utility for the alternative is in the `(Intercept)` column. This is missing for biking because we normalized utility such that this intercept is 0.
- When the cost of the alternative increases by $1, the increase in the student's utility for the alternative is in the `cost` column. This coefficient is the same for all alternatives.
- When the time of the alternative increases by 1 minute, the increase in the student's utility for the alternative is in the `time` column. 

### ii

```{r}
multinomial_cost_coefficient <- multinomial_coefficients[["cost"]]

multinomial_predictions <- predict(multinomial_model, indexed)

commute_multinomial %>%
    select(id, cost.car) %>%
    mutate(
        car_probability = multinomial_predictions[, "car"],
        car = multinomial_cost_coefficient * cost.car * (1 - car_probability),
        # cross elasicities are the same for all other alternatives
        `bike, bus, and walk` =
            -multinomial_cost_coefficient * cost.car * car_probability
    ) %>%
    select(id, car, `bike, bus, and walk`) %>%
    pivot_longer(
        c(car, `bike, bus, and walk`),
        names_to = "Alternatives",
        values_to = "elasticity"
    ) %>%
    group_by(Alternatives) %>%
    summarize(get_quartiles_and_mean(elasticity)) %>%
    pander()
```

The cross-elasticities for substituting driving with biking, taking the bus, or walking are all the same.
The cross-elasticity only depends on the characteristics of the driving alternative.
This is called proportional substitution.
The independence of irrelevant alternatives (IIA) assumption for logistic models implies this result.

## b

### i 

The interpretations of these coefficients are the same as above.

```{r}
marital_status_coefficients <-
    commute_multinomial %>%
    group_by(marital_status) %>%
    summarize(
        # get the coefficients as a tibble
        mlogit(
            multinomial_formula,
            add_indices(cur_data())
        ) %>%
            coef() %>%
            get_coefficients_table(),
        .groups = "drop"
    )

pander(marital_status_coefficients)
```

### ii

I can't use this data to compare the marginal utility of income between the two groups.
Within one regression, the marginal utility of income is the negative of the cost parameter, because costs subtract from income.
However, I used the variance of the error term to scale utility, and this variance might differ between the two groups.

### iii

By the implicit function theorem,

$$ \frac{∂C_{nj}}{∂T_{nj}} = -\frac{β_j}{β_1} $$

This is in dollars per minute.
To get dollars per hour, I will multiply by 60.

```{r}
marital_status_coefficients %>%
    mutate(value = -60 * time / cost) %>%
    select(Alternative, marital_status, value) %>%
    # compare side by side
    pivot_wider(names_from = "marital_status", values_from = "value") %>%
    pander()
```

For all four alternatives, on average, married students have a higher willingness to pay to avoid an hour of commute time.
Perhaps married students value the alternative, time with their family, more.
Or perhaps married students have more responsibilities at home.

## c

### i

```{r}
reduced_time_data <- add_indices(
    commute_multinomial %>%
        # reduce only bus commute times
        mutate(time.bus = time.bus * 0.8)
)

get_expectations <- function(predictions) {
    # Each column is an alternative. Sum columns
    expectations <- apply(predictions, 2, sum)
    tibble(
        Alternative = names(expectations),
        expectation = expectations
    )
}

bind_rows(
    get_expectations(multinomial_predictions) %>%
        mutate(scenario = "Original Time"),
    get_expectations(predict(multinomial_model, reduced_time_data)) %>%
        mutate(scenario = "Reduced Time")
) %>%
    pivot_wider(names_from = scenario, values_from = expectation) %>%
    mutate(Difference = `Reduced Time` - `Original Time`) %>%
    pander()
```

The change in the number of students who choose each alternative is in the `Difference` column.
Several more students choose to take the bus, and most of those students formerly drove.
However, there a few students who formerly walked or biked, and now take the bus.
This might increase the carbon footprint of the policy somewhat.

### ii

The costs are small relative to any one student's income, so I can assume the marginal utility of income is constant for any one student.
However, if one only uses utility to represent individual choices, utility is not interpersonally comparable.
Measuring consumer surplus in dollars instead appeals to the compensation principle: how much students are willing to pay for the lower commute times.
This weighs wealthier students more, who have a higher ability to pay, and thus willingness to pay.

```{r}
tibble(
    `Original Time` =
        sum(logsum(multinomial_model, data = indexed)) /
            -multinomial_cost_coefficient,
    `Reduced Time` =
        sum(logsum(multinomial_model, data = reduced_time_data)) /
            -multinomial_cost_coefficient
) %>%
    mutate(Difference = `Reduced Time` - `Original Time`) %>%
    pander()
```

The increase in consumer surplus is in the `Difference` column.
